MLH1 Results Outline

### 2020-04-07 doc/Results_v2.rmd
updating values

#Fig1_v1_strain + female.specific.plot - male.plot + plot_layout(ncol=1)
Fig1_v1_strain / (female.specific.plot + male.plot )

#(p1 | p2 | p3)/p4

@@

To situate variation in recombination rate within an evolutionary framework, we
We applied fit a series of models including subspecies, strain, and sex, to to our dataset of mean MLH1 foci counts from `r length(DF.HetC.MixedModel.HQ$mean_co)` mice to fit the patterns of variation across sex, subspecies and strain into an evolutionary framework. We began with a full mixed model (M1, see Methods), which showed that strain 
(random effect p < `r MLH1.MMlmer_results$rand.pval`   **someSmallNumber**), sex (p = `r MLH1.MMlmer_results$sex.results` -- 3.9710^{-5}), subspecies (p=`r MLH1.MMlmer_results$interaction.pval`--  1.7210^{-4}), and subspecies sex (p = `r MLH1.MMlmer_results$interaction.pval` -- 3.110^{-5}) each significantly affect recombination rate.

<for values> The linear model results confirm the qualitative observations of sex-specific evolution and significant difference in the degree of sexual dimorphism in genome-wide recombination rates. The subspecies fixed effects tested in M2, were either weak or not significant (glm; p=`r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][2]` and p= `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][3]` for _M. m. musculus_ and _M. m. molossinus_ respectively), indicating that most of the variance is due to strain effects.

As such, we focus on results from M3. The two most significant strain effects were _domesticus^G^_ and _musculus^MSM^_ (p= `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` and p= `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]` for _domesticus^G^_ and _molossinus^MSM^_ respectively). In addition to significant strain effects for _molossinus^MSM^_, (glm; p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][5]` and p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]`)  both _musculus^PWD^_ and _musculus^MSM^_ had significant strain by sex effects (glm; p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`
and p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`) _musculus^PWD^_ by male and _musculus^MSM^_ by male. Combined, these results demonstrate that i) the significant sex-specific evolution in MSM and PWD and ii) the larger magnitude of heterochiasmy in _domesticus^G^_.

We next fit general linear models separately for 192 males and 144 females (M4). In the male dataset, three strains significantly affect recombination rate: molossinus^MSM^ (p = 2.2310^{-14}; effect = X MLH1 foci), All of the effects within the mixed model M1 were highly significant (LRT: subsp p= 1.7210^{-4}, sex p = 3.9710^{-5}, subspecies * sex p = 3.110^{-5}) and random LTR; strain p= 0 ).

**M1. Mixed Model** $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex + rand(strain) + \varepsilon $$

**M2. Linear Model** $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex * strain + \varepsilon$$

**M3. Linear Model** $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ sex * strain + \varepsilon$$

**M4. Linear Model** $$sex\ specific\ mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * strain + \varepsilon$$

The linear model results confirm the qualitative observations of sex-specific evolution and significant difference in the degree of sexual dimorphism in genome wide recombination rates. The subspecies fixed effects tested in M2, were either weak or not significant (glm; p=0.22 and p= 0.09 for *M. m. musculus* and *M. m. molossinus* respectively), indicating that most of the variance is due to strain effects. As such, we focus on results from M3. The two most significant strain effects were domesticusG and musculus^MSM^ (p= 1.0410^{-6} and p= 3.9910^{-6} for domesticusG and molossinusMSM respectively). In addition to significant strain effects for molossinus^MSM^, (glm; p = 0.06 and p = 3.9910^{-6}) both musculusPWD and musculus^MSM^ had significant strain by sex effects (glm; p = 3.8610^{-4} and p = 1.2610^{-4}) musculus^PWD^ by male and musculusMSM by male. Combined, these results demonstrate that i) the significant sex-specific evolution in MSM and PWD and ii) the larger magnitude of heterochiasmy in domesticusG.

A linear model was run on the sex specific data sets (M4), with `r length(fomo.male_mouse.av$mouse)` -- 192 male and `r length(female_mouse.avs_for.analysis$mouse)` -- 144 female mean MLH1 foci per cell. There were three significant strain effects in the male specific data; musculus^PWD^ ((glm; p = 7.3710^{-10} -- `r sum_MLH1.male.specific.glm.M2$coefficients[5,][4]`; effect = ~ 5 X foci), and musculus^SKIVE^ (glm; p = 0.01; `r sum_MLH1.male.specific.glm.M2$coefficients[8,][4]` -- effect = ~7 foci), and  _molossinus^MSM^_ (glm; p=`r sum_MLH1.male.specific.glm.M2$coefficients[6,][4]`; effect ~ 2 foci).

These three strains point to rapid evolution in recombination rate in spermatocytes; we subsequently refer to them collectively as "high-recombination" strains. Analysis of the female dataset points to four strains with significant effects on recombination rate:  and molossinus^MSM^ (glm; p=2.2310^{-14} with effect sizes ranging from 5, 7, and 2 foci respectively.  Given these results, theses strains are classified as 'high recombining' strains and point to rapid evolution in the recombiantion rate for spermatocytes.
While the female specific linear model has many more had significant strain effects; domesticus^G^ (glm; p = 2.510^{-6}), molossinus^MSM^ (p = 6.2410^{-6}), domesticus^LEW^ (glm; p = 0.01), and domesticusPWD musculusPWD (glm; p=0.02). Strain effect sizes in females are modest in magnitude (ranging from 1 to 4 foci) compared to those in males., domesticus^MSM^ (glm; p=6.2410^{-6}), domesticus^MOLF^(glm; p=0.08), and domesticus^KAZ^ (glm; p=0.1), the effect sizes have a smaller range of (4 to 1), indicating significant but small strain specific effects on mean MLH1 foci per cell.
Together, these results demonstrate heritable differences in the genome-wide recombination rate evolving in a highly sex-specific manner over short evolutionary timescales.

@@

We applied a series of models to our dataset of mean MLH1 foci counts from `r length(DF.HetC.MixedModel.HQ$mean_co)` mice to fit the patterns of variation across sex, subspecies and strain into an evolutionary framework. All of the effects within the mixed model M1 were highly significant (LRT: subsp p= `r MLH1.MMlmer_results$sub.sp`, sex p = `r MLH1.MMlmer_results$sex.results`, subspecies * sex p = `r MLH1.MMlmer_results$interaction.pval`) and random LTR; strain p= `r MLH1.MMlmer_results$rand.pval`).

@@

The linear model results confirm the qualitative observations of sex-specific evolution and significant difference in the degree of sexual dimorphism in genome-wide recombination rates. The subspecies fixed effects tested in M2, were either weak or not significant (glm; p=`r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][2]` and p= `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][3]` for _M. m. musculus_ and _M. m. molossinus_ respectively), indicating that most of the variance is due to strain effects. As such, we focus on results from M3. The two most significant strain effects were _domesticus^G^_ and _musculus^MSM^_ (p= `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` and p= `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]` for _domesticus^G^_ and _molossinus^MSM^_ respectively). In addition to significant strain effects for _molossinus^MSM^_, (glm; p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][5]` and p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]`)  both _musculus^PWD^_ and _musculus^MSM^_ had significant strain by sex effects (glm; p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`
and p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`) _musculus^PWD^_ by male and _musculus^MSM^_ by male. Combined, these results demonstrate that i) the significant sex-specific evolution in MSM and PWD and ii) the larger magnitude of heterochiasmy in _domesticus^G^_.

A linear model was run on the sex specific data sets (M4), with `r length(fomo.male_mouse.av$mouse)` male and `r length(female_mouse.avs_for.analysis$mouse)` female mean MLH1 foci per cell. There were three significant strain effects in the male specific data; _musculus^PWD^_ (glm; p=`r sum_MLH1.male.specific.glm.M2$coefficients[5,][4]`), _musculus^SKIVE^_ (glm; p=`r sum_MLH1.male.specific.glm.M2$coefficients[8,][4]`) and _molossinus^MSM^_ (glm; p=`r sum_MLH1.male.specific.glm.M2$coefficients[6,][4]` with effect sizes ranging from 5, 7, and 2 foci respectively. Given these results, theses strains are classified as 'high recombining' strains and point to rapid evolution in the recombiantion rate for spermatocytes.

While the female specific linear model has many more had significant strain effects; _domesticus^G^_ (glm; p=`r sum_MLH1.female.spc.glm.M2$coefficients[2,][4]`), _domesticus^LEW^_ (glm; p=`r sum_MLH1.female.spc.glm.M2$coefficients[3,][4]`), _domesticus^PWD^_ (glm; p=`r sum_MLH1.female.spc.glm.M2$coefficients[4,][4]`), _domesticus^MSM^_ (glm; p=`r sum_MLH1.female.spc.glm.M2$coefficients[5,][4]`), _domesticus^MOLF^_(glm; p=`r sum_MLH1.female.spc.glm.M2$coefficients[6,][4]`), and _domesticus^KAZ^_(glm; p=`r sum_MLH1.female.spc.glm.M2$coefficients[8,][4]`), the effect sizes have a smaller range of (4 to 1), indicating significant but small strain specific effects on mean MLH1 foci per cell.

<domesticus^G^ had significant strain (p= `r summ.MLH1.M3.strain.fixed$coefficients[4,][3]`)
and sex by strain effects (p =`r  summ.MLH1.M3.strain.fixed$coefficients[,4][10]`>

M2(glm; PWD*male p= `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][14]` and MSM-male p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][15]`)>
<Additionally, _domesticus^G^_ had significant strain and sex by strain effects (p =`r  summ.MLH1.M3.strain.fixed$coefficients[,4][3]` and `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]` for strain and strain by sex effect). These results confirm the qualtative pattern of the _domesticus^G^_ strain having the largest degree of sexual dimorphism (heterochiasmy) in the dataset.

In order to test the effects of subspecies, sex and strain on the mean of MLH1 counts per cell, we fit `r length(DF.HetC.MixedModel.HQ$mean_co )` mouse averages of MLH1 foci per cell to a series of mixed and linear models. 
(*note on sampling /sample numbers, the numbers of mice across strains, is shown in supplemental table )
Check the DF of the number cited above
This is the break down of mice used within the Mixed model
`r kable(table( DF.HetC.MixedModel.HQ$sex, DF.HetC.MixedModel.HQ$strain))`
add number of cells
`r kable(table( all.cell.MM.HQ$sex, all.cell.MM.HQ$strain))`
 $$mouse \ av \ meteric ~=~ subsp * sex  + rand(1|strain) + \varepsilon$$

- only Mus musculus strains with sex matched observations  
- Male gwRR have been adjusted for the PAR
- Q5 excluded
- All the effects are un-ordered factors  
- male ages within 5 - 20 weeks
- Molossinus is coded as a subsp level
also g strain effect/ dom^G^ strain effect 
`r summ.MLH1.M2.glm.strain.fixed.full$coefficients[12,][4]` 
In both models strain by sex interaction effects were significant for (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]` and `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]` for the second mixed model) for _musculus^PWD^_ and _molossinus^MSM^_  respectively. 
-->

>>>>

#Fig1_v1_strain + female.specific.plot - male.plot + plot_layout(ncol=1)
#Fig1_v1_strain / (female.specific.plot + male.plot )

#(p1 | p2 | p3)/p4
```

@@

@@

@@

To situate variation in recombination rate within an evolutionary framework, we fit a series of models including subspecies, strain, and sex, to mean MLH1 foci counts from `r length(DF.HetC.MixedModel.HQ$mean_co)` mice.

We began with a full mixed model (M1, see Methods), which showed that strain 
(random effect p < `r MLH1.MMlmer_results$rand.pval`   **someSmallNumber**), sex (p = `r MLH1.MMlmer_results$sex.results` -- 3.9710^{-5}), subspecies (p=`r MLH1.MMlmer_results$sub.sp`--  1.7210^{-4}), and subspecies sex (p = `r MLH1.MMlmer_results$interaction.pval` -- 3.110^{-5}) each significantly affect recombination rate.

We next fit general linear models separately for 192--`r length(fomo.male_mouse.av$mouse)` males and `r length(female_mouse.avs_for.analysis$mouse)`--144 females (M4). In the male dataset, three strains significantly affect recombination rate: musculus^PWD^ ((glm; p = 7.3710^{-10} -- `r sum_MLH1.male.specific.glm.M2$coefficients[5,][4]`; effect = ~ 5 X foci), and musculus^SKIVE^ (glm; p = 0.01; `r sum_MLH1.male.specific.glm.M2$coefficients[8,][4]` -- effect = ~7 foci), and  _molossinus^MSM^_ (glm; p=`r sum_MLH1.male.specific.glm.M2$coefficients[6,][4]`; effect ~ 2 foci). These three strains point to rapid evolution in recombination rate in spermatocytes; we subsequently refer to them collectively as "high-recombination" strains.  Analysis of the female dataset points to four strains with significant effects on recombination rate: domesticus^G^ (p = 2.510^{-6} -- `r sum_MLH1.female.spc.glm.M2$coefficients[2,][4]`), molossinus^MSM^ (p = 6.2410^{-6} -- `r sum_MLH1.female.spc.glm.M2$coefficients[5,][4]`), domesticus^LEW^ (p = 0.01 -- `r sum_MLH1.female.spc.glm.M2$coefficients[3,][4]`), and musculus^PWD^ (p=0.02  -- `r sum_MLH1.female.spc.glm.M2$coefficients[4,][4]`). Strain effect sizes in females are modest in magnitude (ranging from 1 to 4 foci) compared to those in males.

Together, these results demonstrate heritable differences in the genome-wide recombination rate evolving in a highly sex-specific manner over short evolutionary timescales.
